Outline paper - Multiple imputation for statistical disclosure control: Creating synthetic datasets with *mice*

### 2020-11-29 mice_synthesizing/paper.Rmd

Open science, including open data, has been marked as the future of science [@gewin_data_2016], and the advantages of publicly available research data are numerous [@molloy_open_2011; @walport_brest_sharing_2011]. Because collecting research data requires an enormous investment both in terms of time and monetary resources, openly accessible research data bears the potential of increasing the scientific returns. Additionally, the fact that public funds are used for data collection results in increased demands for the collected data. Nevertheless, the possibilities to share research data directly are often very limited due to restrictions with regard to privacy and confidentiality. Simply anonimizing the data is not enough to fulfil these requirements [@ohm_broken_2009]. Over the years, several other techniques have been used to increase the confidentiality of the data, such as categorizing continuous variables, top coding values above a certain threshold or adding random noise [@drechsler_synthetic_2011]. However, these methods may distort relationships between variables, thereby drastically reducing the data quality and the usability for further research.

An alternative solution has been proposed separately by @rubin_statistical_disclosure_1993 and @little_statistical_1993. Although their approaches differed to some extent, the general idea is to use actually observed data to generated multiply imputed synthetic datasets that can be freely disclosed. Based on some imputation model, values from the actually observed data are replaced by multiple draws from the posterior predictive distribution of the observed data. Using this approach, the researcher could replace the observed data set as a whole with multiple synthetic versions. Alternatively, the researcher could opt to only replace a subset of the observed data. The values that are to be replaced can be complete variables that could be compared with publicly available data sets or registers, such as addresses, or can be certain values that bear a high risk of being disclosive, say, income values above a certain threshold.

Conceptually, the synthetic data framework is based upon the building blocks of multiple imputation of missing data, as proposed by @rubin_multiple_1987. Instead of replacing just the missing values with multiple draws from the posterior predictive distribution, one could easily *overimpute* any observed sensitive values. Similarly to multiple imputation of missing data, the multiple synthetic datasets allow for correct statistical inferences. The results of the multiply synthesized data sets should be pooled into a single inference, so that the researcher can draw valid inference from a single pooled analysis. To that respect, the variance should reflect the added variability that is induced by the imputation procedure.

Potentially, this approach could fulfill the needs for openly accessibly data, without running into barriers with regard to privacy and confidentiality constraints. However, there is no such thing as a free lunch: data collectors have to put effort in creating high-quality synthetic data. Also, the quality of the synthetic data are highly dependent on the imputation models, and using flawed models to generate synthetic data might bias subsequent analyses. Conversely, if the models used to create the synthetic data are able to preserve the relationships between the variables as in the original data, the synthetic data can be nearly as informative as the observed data. Thus, to fully exploit the benefits of synthetic data, the effort to actually create these datasets should be kept at a minimum.

To minimize additional effort of creating synthetic datasets on behalf of the researcher, software aimed at multiple imputation of missing data can be employed. Especially if researchers used this software at an earlier stage in the research process, or acquired familiarity with it during earlier projects, the additional burden of creating synthetic datasets is relatively small. The R-package `mice` [@mice] implements multiple imputation of missing data in a straightforward and user-friendly manner. However, the functionality of `mice` is not restricted to the imputation of missing data, but allows to impute any value in the data. Thus, `mice` can be utilized for the creation of multiply imputed synthetic datasets.

After creating the multiply imputed synthetic data sets, the goal is to obtain valid statistical inferences from them. In the missing data framework, this is done by performing statistical analyses on all imputed data sets, and pooling the results of the analyses according to Rubin's rules [-@rubin_multiple_1987]. In the synthetic data framework, the same procedure is followed. However, valid statistical results can only be obtained under the premise that the statistical properties of the analyses are correct. Analyzing synthetic data specically requires appropriate variance estimates. Notably, the variance estimator that is suitable for analyzing multiply imputed datasets for missing data is inappropriate when multiply imputed synthetic datasets are analyzed. As a result, multiple variance estimators have been proposed to analyze multiply imputed synthetic datasets correctly, that differ according to whether the data consists entirely of synthetic values, or only a subset of the data is synthetic.

@@

The `mice` package has been developed originally for multiple imputation of missing data. In this context, the aim is to replace missing values due to nonresponse by plausible values from the posterior predictive distribution of the variable containing the missing values. Doing so, `mice` makes use of fully conditional specification (FCS), which breaks down the multivariate distribution of the data $\textbf{Y} = (\textbf{Y}_{obs}, \textbf{Y}_{mis})$ into $j = 1, 2, \dots, k$ univariate conditional densities, with $k$ the number of variables in the data. Using FCS, for every variable containing missing values a model is constructed, to impute the missing values $Y_{j, mis}$ with draws from the posterior predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ on a variable by variable basis. Note that the predictor matrix $Y_{-j}$ may contain yet imputed values from an earlier imputation step, and thus will be updated after every iteration. This procedure is applied $m$ times, resulting in $m$ completed datasets $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$, with $\textbf{D}^{(l)} = (\textbf{Y}_{obs}, Y^{(l)}_{mis}$. __MAYBE ADD AN EXAMPLE OF HOW TO DO IT IN `mice`__

@@

This approach can be straightforwardly extended to a synthetic data framework. However, rather than imputing missing data, we overimpute actually observed data, that is, we replace observed values by multiple imputations. For simplicity, assume that the data is completely observed (i.e., $\textbf{Y} = \textbf{Y}_{obs}$). Following the notation of @reiter_raghunathan_multiple_2007, let $Z_j = 1$ if any of the values of unit $j$ are to be replaced by imputations, and $Z_j = 0$ otherwise, with $\textbf{Z} = (Z_1, Z_2, \dots, Z_n)$. Accordingly, the data consists of values that are to be replaced and values that are to be kept (i.e., $\textbf{Y}^{(l)} = (\textbf{Y}^{(l)}_{rep}, \textbf{Y}_{nrep})$. Note the superscript $^{(l)}$ indicating the $l$th synthetic data set. Now, instead of imputing $\textbf{Y}^{(l)}_{mis}$ with draws from the posterior predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ as in the missing data case, we impute $\textbf{Y}^{(l)}_{rep}$ from the posterior distribution of $(Y_{rep} | \textbf{Y}, \textbf{Z}, \theta)$. This process results in $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$, with $m$ the number of synthetic datasets.

By default, `mice` imputes all values that are coded as missing in the dataset. Using the `where` parameter within the mice function, one can choose to impute any value in the dataset. As input, the `where` parameter requires a matrix of the same dimensions of the data, (i.e., a $n \times k$ matrix, with $n$ the number of observations in the collected data) containing logical values.
<!--

>>>>

Open science, including open data, has been marked as the future of science [@gewin_data_2016], and the advantages of publicly available research data are numerous [@molloy_open_2011; @walport_brest_sharing_2011]. Because collecting research data requires an enormous investment both in terms of time and monetary resources, openly accessible research data bears the potential of increasing the scientific returns. Additionally, the fact that public funds are used for data collection results in increasing demand for the collected data. Nevertheless, the possibilities to share research data directly are often very limited due to restrictions with regard to privacy and confidentiality. Simply anonimizing the data is not enough to fulfil these requirements [@ohm_broken_2009]. Over the years, several other techniques have been used to increase the confidentiality of the data, such as categorizing continuous variables, top coding values above a certain threshold or adding random noise [@drechsler_synthetic_2011]. However, these methods may distort relationships between variables, thereby drastically reducing the data quality and the usability for further research.

An alternative solution has been proposed separately by @rubin_statistical_disclosure_1993 and @little_statistical_1993. Although their approaches differed to some extent, the general idea is to use actually observed data to generate multiply imputed synthetic datasets that can be freely disclosed. Based on some imputation model, values from the actually observed data are replaced by multiple draws from the posterior predictive distribution of the observed data. Using this approach, the researcher could replace the observed data set as a whole with multiple synthetic versions. Alternatively, the researcher could opt to only replace a subset of the observed data. The values that are to be replaced can be complete variables that could be compared with publicly available data sets or registers, such as addresses, or can be certain values that bear a high risk of being disclosive, say, income values above a certain threshold.

Conceptually, the synthetic data framework is based upon the building blocks of multiple imputation of missing data, as proposed by @rubin_multiple_1987. Instead of replacing just the missing values with multiple draws from the posterior predictive distribution, one could easily *overimpute* any observed sensitive values. Similarly to multiple imputation of missing data, the multiple synthetic datasets allow for correct statistical inferences. The results of the multiple synthetic data sets should be pooled into a single inference, so that the researcher can draw valid conclusions from the pooled results. To that respect, the variance should reflect the added variability that is induced by the imputation procedure.

Potentially, this approach could fulfil the needs for openly accessibly data, without running into barriers with regard to privacy and confidentiality constraints. However, there is no such thing as a free lunch: data collectors have to put effort in creating high-quality synthetic data. Also, the quality of the synthetic data is highly dependent on the imputation models, and using flawed models to generate synthetic data might bias subsequent analyses. Conversely, if the models used to create the synthetic data are able to preserve the relationships between the variables as in the original data, the synthetic data can be nearly as informative as the observed data. Thus, to fully exploit the benefits of synthetic data, the effort to actually create these high-quality data sets should be kept at a minimum.

To minimize additional effort of creating synthetic datasets on behalf of the researcher, software aimed at multiple imputation of missing data can be employed. Especially if researchers used this software at an earlier stage in the research process, or acquired familiarity with it during earlier projects, the additional burden of creating synthetic datasets is relatively small. The R-package `mice` [@mice] implements multiple imputation of missing data in a straightforward and user-friendly manner. However, the functionality of `mice` is not restricted to the imputation of missing data, but allows to impute any value in the data. Consequently, `mice` can be utilized for the creation of multiply imputed synthetic datasets.

After creating the multiply imputed synthetic data sets, the goal is to obtain valid statistical inferences. In the missing data framework, this is done by performing statistical analyses on all imputed data sets, and pooling the results of the analyses according to Rubin's rules [-@rubin_multiple_1987]. In the synthetic data framework, the same procedure is followed. However, valid statistical results can only be obtained under the premise that the statistical properties of the analyses are correct. Analyzing synthetic data specifically requires appropriate variance estimates. Notably, the variance estimator that is suitable for analyzing multiply imputed datasets for missing data is inappropriate when multiply imputed synthetic datasets are analyzed. As a result, multiple variance estimators have been proposed to make correct inferences from multiply imputed synthetic datasets. These estimators differ according to whether the data consists entirely of synthetic values, or only a subset of the data is synthetic.

@@

The `mice` package has been developed originally for multiple imputation of missing data. In this context, the aim is to replace missing values due to nonresponse by plausible values from the posterior predictive distribution of the variable containing the missings. Doing so, `mice` makes use of fully conditional specification (FCS), which breaks down the multivariate distribution of the data $\textbf{Y} = (\textbf{Y}_{obs}, \textbf{Y}_{mis})$ into $j = 1, 2, \dots, k$ univariate conditional densities, with $k$ the number of variables in the data. Using FCS, for every variable containing missing values a model is constructed, to impute the missing values $Y_{j, mis}$ with draws from the posterior predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ on a variable by variable basis. Note that the predictor matrix $Y_{-j}$ may contain yet imputed values from an earlier imputation step, and thus will be updated after every iteration. This procedure is applied $m$ times, resulting in $m$ completed datasets $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$, with $\textbf{D}^{(l)} = (\textbf{Y}_{obs}, Y^{(l)}_{mis})$. __MAYBE ADD AN EXAMPLE OF HOW TO DO IT IN `mice`__

@@

This approach can be straightforwardly extended to a synthetic data framework. However, rather than imputing missing data, we overimpute actually observed data, that is, we replace observed values by multiple imputations. For simplicity, assume that the data is completely observed (i.e., $\textbf{Y} = \textbf{Y}_{obs}$). Following the notation of @reiter_raghunathan_multiple_2007, let $Z_j = 1$ if any of the values of unit $j$ are to be replaced by imputations, and $Z_j = 0$ otherwise, with $\textbf{Z} = (Z_1, Z_2, \dots, Z_n)$. Accordingly, the data consists of values that are to be replaced and values that are to be kept (i.e., $\textbf{Y}^{(l)} = (\textbf{Y}^{(l)}_{rep}, \textbf{Y}_{nrep})$. Note the superscript $^{(l)}$ indicating the $l$th synthetic data set, with $l = 1, 2, \dots, m$. Now, instead of imputing $\textbf{Y}^{(l)}_{mis}$ with draws from the posterior predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ as in the missing data case, we impute $\textbf{Y}^{(l)}_{rep}$ from the posterior distribution of $(Y_{rep} | \textbf{Y}, \textbf{Z}, \theta)$. This process results in the synthetic data $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$.

By default, `mice` imputes all values that are coded as missing in the dataset. Using the `where` parameter within the mice function, one can choose to impute any value in the dataset. As input, the `where` parameter requires a matrix of the same dimensions of the data, (i.e., a $n \times k$ matrix, with $n$ the number of observations in the collected data) containing logical values.
